---
title: "Sleep Disruption as an Endophenotype of Autism: Analysis Report"
author: "Mailton Vasconcelos"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(knitr)
library(kableExtra)

# Source cleaning pipeline
source("AAB_cleaning_pipeline.R")
```

# Sample Description

## Participant characteristics
<!-- Table 1: N by participant type, age (mean/SD), sex, site -->

## Missingness summary
<!-- Overall missingness on key variables by participant group -->

---

# Block 1: Endophenotype Testing

## Step 1a: Candidate Endophenotypes — Sleep Traits (CSHQ)

### Descriptive statistics
<!-- CSHQ total and subscale means, SDs, range by participant group -->

### Distribution checks
<!-- Histograms, QQ plots, Shapiro-Wilk normality tests -->

### Missingness on sleep variables
<!-- % missing by participant type; pattern analysis -->

## Step 1b: Missing Data and Power Analysis

### Missingness patterns
<!-- MAR/MCAR assessment; missingness by group and variable -->

### Multiple imputation
<!-- mice() imputation — methods, m, iterations -->

### Power analysis
<!-- Simulation or analytical power per endophenotype criterion -->

## Step 2: Association with Autism Diagnosis

### Model 1 — Demographics only
<!-- lmer: CSHQ ~ diagnosis_group + sex + age + site + (1|family_id) -->

### Model 2 — Biological covariates added
<!-- Adds tanner, melatonin_use, SES -->

### Model 3 — Full model
<!-- Adds vabs_int_raw, vabs_ext_raw, vabs_comp_standard -->

### Proband severity model
<!-- lmer: CSHQ ~ ados2_comparisonscore + covariates + (1|family_id) -->

### Summary
<!-- Effect sizes, confidence intervals, ICC -->

## Step 3: Heritability Estimation

### Variance components model
<!-- Narrow-sense h² from pedigree-based kinship matrix -->
<!-- Tools: SOLAR or R gap package -->

### Results
<!-- h² estimates with SE and 95% CI per CSHQ subscale -->

## Step 4: Genetic Correlation Analysis

### Co-segregation analysis
<!-- Genetic correlation between sleep traits and autism liability -->
<!-- Tools: SOLAR-Eclipse or R heritability/gap packages -->

### Results
<!-- Genetic vs environmental decomposition of phenotypic correlation -->

## Step 5: Intermediate Expression in Relatives

### Comparison: unaffected relatives vs controls
<!-- Mixed models: sleep traits in siblings vs controls -->

### Results
<!-- Do siblings show intermediate trait expression? -->

## Step 6: Genetic Mediation Analysis

### Mendelian Randomisation
<!-- IVW, MR-Egger, Weighted Median -->
<!-- Tools: TwoSampleMR, MendelianRandomization -->

### Pleiotropy tests
<!-- MR-Egger intercept; sensitivity analyses -->

### Results
<!-- Causal vs pleiotropic pathways -->

---

# Block 2: Latent Dimensions and Prioritisation

## Step 7: Latent Structure Analysis

### Factor analysis
<!-- ML factor analysis on CSHQ subscales; varimax rotation -->
<!-- Parallel analysis for factor number -->

### Results
<!-- Factor loadings, variance explained, factor interpretation -->

## Step 8a: Endophenotype Ranking (ERV)

### ERV calculation
<!-- Endophenotype Ranking Variable per sleep trait -->
<!-- Bootstrap stability -->

### Results
<!-- Ranked table of CSHQ subscales by ERV -->

## Step 8b: Endophenotype Quality Index (EQI)

### EQI scoring
<!-- Liu & Gershon 2024 framework — 12 criteria -->
<!-- Reliability, precision, throughput, cost, generalisability -->

### Combined ERV + EQI ranking
<!-- Composite prioritisation table -->

---

# Block 3: Multi-Dimensional Integration

## Step 9: Structural Equation Modelling

### Model specification
<!-- Latent sleep factors + PGS + environmental variables + autism severity -->

### Results
<!-- Path coefficients, model fit indices -->

### Bayesian network modelling
<!-- If applicable -->

---

# Block 4: Secondary Analyses

## Step 10: Secondary Endophenotypes

### Cognition
<!-- Same pipeline applied to cognitive variables -->

### Sensory profiles
<!-- Same pipeline applied to SSP -->

### Comparative ranking
<!-- Sleep vs cognition vs sensory as endophenotypes -->

---

# Summary and Conclusions

## Key findings per endophenotype criterion
<!-- Table summarising evidence across Steps 2–6 -->

## Limitations
<!-- Chronotype absent, cross-sectional, instrument differences, etc. -->

## Implications for endophenotype status of sleep disruption

---

# Supplementary Materials

## Sensitivity analyses
## Imputation diagnostics
## Full model outputs
## EQI scoring sheets

---

# References

---

# Session Information
```{r session-info}
sessionInfo()
```