---
title: "Missingness Inspection and Step 2 Modelling Decisions"
author: "Mailton Vasconcelos"
date: "2026-03-28"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: kate
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
```

---

# Overview

This report documents the pre-modelling missingness inspection for the FamilySleeps
endophenotype analysis (Step 2: Association with Autism Diagnosis). The analytic
sample comprises autistic children (ASD), non-autistic siblings (Sibling), and
unrelated non-autistic children (Comparison) from the Australian Autism Biobank (AAB).
All analyses use the cleaned dataset produced by `Get-vars-new_20260328v02.R`.

---

# 1. Analytic sample

```{r sample-flow}
tribble(
  ~Step, ~n, ~Loss, ~Note,
  "Participant types 1 / 3 / 4 (ASD, Sibling, Comparison)", 1564, "—", "",
  "Confirmed diagnosis_group (excludes unconfirmed probands & ASD-Q)", 1372, "192", "Unconfirmed ASD diagnosis or ASD-Q",
  "CSHQ total prorated available", 1211, "161", "Structural CSHQ missingness",
  "Complete on age, sex, melatonin_ever, family_id", 1124, "87",  "Age / sex missingness in siblings and Comparison",
  "Complete on ethnicity_collapsed (full model)", 1101, "23",  "Comparison children missing family-level ethnicity"
) %>%
  kable(col.names = c("Step", "n", "Loss", "Note"),
        align = c("l","r","r","l")) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE) %>%
  row_spec(5, bold = TRUE)
```

**Final analytic N: 1,101** (minimal model without ethnicity: n = 1,124).

---

# 2. Overall missingness

```{r overall-miss}
tribble(
  ~Variable,                  ~pct_miss, ~n_missing, ~n_present,
  "tanner_genital",               76.7,      1199,        365,
  "ses_composite_complete",       35.7,       559,       1005,
  "ses_composite_partial",        25.3,       396,       1168,
  "cshq_total_prorated",          19.6,       307,       1257,
  "diagnosis_group",              12.3,       192,       1372,
  "ethnicity_collapsed",          11.9,       186,       1378,
  "family_id",                     6.8,       107,       1457,
  "sex",                           0.6,        10,       1554,
  "age",                           0.4,         6,       1558,
  "site",                          0.0,         0,       1564,
  "melatonin_ever",                0.0,         0,       1564
) %>%
  kable(col.names = c("Variable", "% Missing", "n Missing", "n Present"),
        align = c("l","r","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE) %>%
  row_spec(c(1,2,3), background = "#fff3cd")
```

Highlighted rows are variables excluded from main models (see Section 5).

---

# 3. Missingness by diagnosis group

```{r miss-by-group}
tribble(
  ~Variable,                ~ASD,  ~Sibling, ~Comparison,
  "cshq_total_prorated",    16.6,     28.4,       27.5,
  "ethnicity_collapsed",    11.4,      4.2,       28.9,
  "tanner_genital",         77.9,     67.8,       82.6,
  "ses_composite_complete", 32.0,     15.7,      100.0,
  "ses_composite_partial",  19.7,      7.7,      100.0,
  "family_id",               9.1,      0.8,        0.0,
  "age",                     0.2,      1.5,        0.0,
  "sex",                     0.4,      1.5,        0.7,
  "melatonin_ever",          0.0,      0.0,        0.0
) %>%
  kable(col.names = c("Variable", "ASD (%)", "Sibling (%)", "Comparison (%)"),
        align = c("l","r","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE)
```

Key patterns: SES is 100% missing for Comparison children (structural); Tanner missingness
is high across all groups; CSHQ missingness is higher in Sibling and Comparison (~28%)
than ASD (~17%), consistent with the FHQ being completed primarily for the proband.

---

# 4. Missingness by site

```{r miss-by-site}
tribble(
  ~Variable,                  ~WA,   ~VIC,  ~NSW,  ~QLD,
  "cshq_total_prorated",      16.4,  21.9,  20.8,  23.2,
  "tanner_genital",           84.1,  84.9,  74.0,  57.3,
  "ses_composite_complete",   42.6,  36.3,  30.1,  27.5,
  "ses_composite_partial",    35.1,  24.3,  14.5,  18.2,
  "family_id",                16.1,   0.8,   0.0,   0.3,
  "pedigree_linked",          13.6,   8.4,   6.0,   6.0,
  "ethnicity_collapsed",      11.5,  17.5,   9.0,  11.6
) %>%
  kable(col.names = c("Variable", "WA (%)", "VIC (%)", "NSW (%)", "QLD (%)"),
        align = c("l","r","r","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE) %>%
  row_spec(c(5,6), background = "#fff3cd")
```

Highlighted rows flag the WA site data quality issue (see Decision 3 below).
Tanner missingness is notably lower in QLD (57.3%) than other sites (~74–85%),
suggesting a site-level protocol difference.

---

# 5. MAR checks — is CSHQ missingness predicted by observed variables?

```{r mar-checks}
tribble(
  ~Test, ~Statistic, ~p, ~Interpretation,
  "Diagnosis group (χ²)", "χ²(3) = 540.19", "< 0.001", "Driven by unconfirmed probands; resolved by diagnosis_group filter",
  "Site (χ²)",            "χ²(3) = 7.82",   "0.050",   "Marginal; site included as covariate in all models",
  "Sex (χ²)",             "χ²(2) = 28.45",  "< 0.001", "Males missing less; attenuates once diagnosis group is controlled",
  "Age (t-test)",         "t(467) = 1.99",  "0.047",   "Missing children ~0.5 yrs younger; age is a model covariate",
  "SES (t-test)",         "t(120) = 2.87",  "0.005",   "Missing children have lower SES; SES excluded from main models"
) %>%
  kable(col.names = c("Test", "Statistic", "p", "Interpretation"),
        align = c("l","l","r","l")) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE)
```

CSHQ missingness is not MCAR. The data are consistent with a **Missing At Random (MAR)**
mechanism — missingness is predicted by observed variables (diagnosis group, site, sex,
age, SES) that are either included as covariates or removed by the analytic filter.
Complete-case analysis is the primary approach; multiple imputation via `mice()` will
be used as a sensitivity analysis (Step 1b).

---

# 6. Descriptive statistics — analytic sample

```{r table1}
tribble(
  ~Group,       ~n,  ~Mean_age, ~SD_age, ~pct_male, ~Mean_CSHQ, ~SD_CSHQ, ~pct_mel,
  "Comparison", 89,   5.8,       3.2,     49.4,       43.3,        7.7,      1.1,
  "Sibling",   182,   8.1,       4.3,     48.4,       45.1,        9.1,      1.1,
  "ASD",       830,   7.5,       3.8,     78.7,       49.8,       10.0,     14.8
) %>%
  kable(col.names = c("Group", "n", "Mean age", "SD", "% Male",
                       "Mean CSHQ", "SD", "% Melatonin"),
        align = c("l","r","r","r","r","r","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE)
```

CSHQ scores increase in the expected direction: Comparison < Sibling < ASD,
providing initial support for intermediate expression in relatives (Step 5).
Melatonin use is strongly differentiated (ASD 14.8% vs 1.1% in other groups),
confirming it as a necessary covariate. Comparison children are on average
younger than ASD and Sibling groups (5.8 vs 7.5–8.1 years).

---

# 7. Modelling decisions

## Decision 1 — Tanner stages excluded from main models {.decision}

Tanner genital staging is missing in 76.7% of the analytic sample. The dominant
reason is `not_applicable_na` (n = 880, assessor-marked), which is age-dependent
and not random. Missingness is also site-patterned (57% QLD vs 84% WA/VIC).
Including Tanner in main models would silently restrict the sample to older children
and introduce site confounding. **Age is the sole developmental proxy in all main
models.** Tanner will be included in a sensitivity analysis restricted to the
Tanner-complete subset (n = 365; primarily QLD).

## Decision 2 — SES excluded from main models {.decision}

SES is derived from the proband FHQ row and propagated via `family_id`. Comparison
children have singleton family IDs with no proband, so SES is 100% missing for
that group. Including SES in the main models would exclude all Comparison children,
eliminating the reference group. **SES (`ses_composite_partial`) will be included
in a sensitivity analysis restricted to ASD + Sibling only**, where coverage is
80.3% and 92.3% respectively.

## Decision 3 — WA site sensitivity analysis {.decision}

`family_id` and `pedigree_linked` missingness is concentrated in WA (16.1% and
13.6% respectively), consistent with the known data entry omission of parent IDs
at that site (104/105 unlinked probands are from WA). This affects family clustering
in mixed-effects models and pedigree completeness for heritability analyses.
**All main models will be re-run excluding WA** to test robustness to this data
quality issue.

---

# 8. Step 2 model specification

**Outcome:** `cshq_total_prorated` (primary); CSHQ subscales (secondary)

**Fixed effects:**
- `diagnosis_group` (Comparison / Sibling / ASD; Comparison as reference)
- `age` (continuous)
- `sex` (Male / Female)
- `melatonin_ever` (0 / 1)
- `ethnicity_collapsed` (Caucasian / Asian / Indigenous-Pacific / Other; Caucasian as reference)
- `site` (WA / VIC / NSW / QLD)

**Random effect:** `(1 | family_id)` — random intercept for family clustering

**Analytic N:** 1,101 (full model); 1,124 (without ethnicity)

**Sensitivity analyses:**

| Sensitivity | Modification | Purpose |
|---|---|---|
| S1 | Exclude WA site | Test robustness to WA data quality |
| S2 | ASD + Sibling only; add `ses_composite_partial` | SES effect |
| S3 | Tanner-complete subset; replace `age` with `age + tanner_genital` | Pubertal adjustment |
| S4 | Multiple imputation via `mice()` | MAR assumption |

---

*Script:* `Get-vars-new_20260328v02.R` | *Missingness inspection:* `1b_Handling-Missing-Data_20260328v01.R`
